Kindly Bent to Free Us

### 2020-02-28 kindly/lang.tex
Adapt the figures to fit better.
cosmetics

\lang is a minimal ML-like language with
let-polymorphism and abstract types. Its type system manages
linearity and borrows using kinds and (kind-) constrained types.
For ease of presentation, we simplify the language.
\begin{itemize}[topsep=0pt]
\item Pattern matching is demonstrated on pairs rather than algebraic
  datatypes.
\item There are separate operators for borrowing and reborrowing (taking
  the borrow of a borrow).
\item Regions are explicit in the code and must be annotated using the
  algorithms presented in \cref{regionannot}.
\item Regions are identified using nesting levels instead of region
  variables.
\end{itemize}

@@

A borrow $\borrow{x}$ is always taken from a variable $x$. The
\emph{borrow annotation} $\BORROW$ indicates whether the borrow is exclusive
($\MBORROW$) or shared ($\IBORROW$). If $x$ is already borrowed,
then we need to use the reborrow expression $\reborrow{x}$.
%
A region $\region{\Sone x\BORROW}{e}$ is annotated with its nesting $n$, the variable $x$ that may be borrowed in $e$, and the kind of borrow $\BORROW$.
%
A match is indexed with a \emph{match specification} $\etransfm$ that indicates
whether the match should work on borrows ($\etransfm=\&^b$) or not ($\etransfm=id$).
%
We consider four primitive operations to manipulate resources:
$\create$, $\observe$, $\update$ and $\destroy$.

@@

In \cref{motivation}, region annotations are optional.
In our formalization, regions must be syntactically explicit and annotated with
a nesting index and a scoped variable.
We now define a best-effort
syntactic code transformation $\RannotT{p}{p'}$ which
annotates programs automatically.
The input $p$ is a program with optional region annotations of the form $\region[]{}{e}$.
The output $p'$ is a program with explicit annotations
of the form $\region{\Sone x\BORROW}{e}$.
We give an informal presentation of our code transformation and defer
the complete definition to \cref{appendix:regionannot}. This code transformation
aims to find, for each borrow $\borrow{x}$, the biggest region satisfying
the following rules:
\begin{enumerate}
\item The region should contain at least $\borrow{x}$.
\item The region should never be larger than the scope of $x$.
\item An exclusive borrow $\borrow[\kaff]{x}$ should never share a region with any other borrow of $x$.
\item A use of $x$ can never be in the region of $\borrow{x}$.
\end{enumerate}
We then start from the leaves and grow the region
associated to each borrow until it can't grow anymore either due to a binding
or to conflicting uses. We demonstrate
on the following program:
%
\[
  \setlength{\jot}{-1pt}
  \RannotT{
\begin{aligned}
  \lam{a}{}\ &
  \letin{x}{\app{f}{\borrow[\kun]{a}}}{}\\
  &g\ (\borrow[\kaff]{x});\\
  &f\ (\borrow[\kun]{x})\ (\borrow[\kun]{x})\\
\end{aligned}
}{
\begin{aligned}
  \lam{a}{}\region[1]{\Sone a\MBORROW}{\ &%
  \letin{x}{\app{f}{\borrow[\kaff]{a}}}{}\\
  &\region[2]{\Sone x\MBORROW}{g\ (\borrow[\kaff]{x})};\\
  &\region[2]{\Sone x\IBORROW}{f\ (\borrow[\kun]{x})\ (\borrow[\kun]{x})}\\
  &\!\!\!}
\end{aligned}
}
\]
%
Since $a$ only has one borrow, its region covers its whole lexical scope.
$x$ has multiple conflicting borrows and requires more consideration.
We place a first
region around the exclusive borrow and its function application,
and a second region around both shared borrows. This placement of region
is optimal: making any region bigger would cause a borrowing
error.
Note that indices of regions are well nested, as the inner region have
higher indices than the outer ones.

Programmers might also want to annotate regions explicitly.
Such annotations are considered as
upper bound on the contained regions.
In the following program, a manual annotation has been inserted to ensure
no borrow enters the reference $r$:
%
\[
  \setlength{\jot}{-1pt}
  \RannotT{
\begin{aligned}
  \letin{&r}{\operatorname{ref}\ 0}{}\\
  \lam{a}{}\ &
  set\ r\ \region[]{}{g\ (\borrow[\kun]{a})};\\
  &f\ (\borrow[\kun]{a})
\end{aligned}
}{
\begin{aligned}
  \letin{&r}{\operatorname{ref}\ 0}{}\\
  \lam{a}{}\ &
  set\ r\ \region[1]{\Sone a\IBORROW}{g\ (\borrow[\kun]{a})};\\
  &\region[1]{\Sone a\IBORROW}{f\ (\borrow[\kun]{a})}
\end{aligned}
}
\]
%
The rules allow merging the two regions around the borrows
$\borrow[\kun]{a}$. However
the explicit annotation indicates that the region should stay around the closure
passed as argument of $set$. This is useful to control captures
by imperative APIs.

This code transformation is purely syntactic and must be used before
typing. It only produces well-nested annotations: if $\region[n]{b}{\dots}$
is nested inside $\region[n']{b'}{\dots}$, then $n > n'$. Furthermore, there
is at most one region per borrow, and exactly one region per exclusive borrow.
In the rest of this article, we consider
that all terms have been annotated with this code transformation and respect
these properties.

@@

We now focus on the essential and novel parts of the type system. A complete
description is available in \cref{appendix:sdtyping}.
% In this section, we focus on chosen parts of our system
% through a set of selected rules.
%
We discuss the following judgments:
\begin{description}
\item[$\inferS{C}{\E}{e}{\tau}$ --]
  Expression $e$ has type $\tau$ in environment $\E$ under constraints $C$.
\item[$\inferSK{C}{\E}{\tau}{k}$ --]
  Type $\tau$ has kind $k$ in environment $\E$ under constraints $C$.
\item[$\entail{D}{C}$ --] Constraint $D$ entails constraint $C$.
\end{description}

@@

\affe uses kinds and constrained types to indicate
linear and affine types.
% Before explaining the details of our type system, we
% give a quick presentation of the kinds and constraints.
A kind $k$ is either a kind variable, $\kvar$, or a constant $\mul_n$.
The quality $\mul$ describes the use pattern of the type:
unrestricted $\kun$, affine $\kaff$, or linear $\klin$. The level $n \in \Nat \cup \{\infty\}$
describes the nested regions in which the value can be used.
Level $0$ refers to the top-level scope outside any region; we often elide it
and write $\kaff$ for $\kaff_0$. Level
$\infty$ refers to an empty region that is infinitely nested.
For instance, the constraint $\Cleq{\kvar}{\kun_\infty}$ indicates that
$\kvar$ must be unrestricted, but can be local to a region.
%
Kinds form a lattice described in \cref{sdtyp:lattice}.
Unrestricted values may be used
where affine ones are expected and affine ones are less restricted
than linear ones as reflected in \TirName{Lat-UAL}.
Values usable at level $n$ may be used at any more deeply
nested level $n'$ as defined in the \TirName{Lat-Level} axioms.
%
Constraints are conjunctions of inequality constraints over this kind
lattice. In most uses, constraints give upper or lower bounds for
kind variables or relate two kind variables.

@@

\begin{figure}[tp]
  \begin{align*}
    \E &::=\ \Eempty \mid \E;B\tag{Environments}\\
    B &::=\ \bnone % \tag{Empty} \\
    \mid \bvar{\tvar}{\kschm} % \tag{Types}
    \mid \bvar{x}{\schm}% \tag{Variables}
    \mid \bvar{\borrow{x}}{\borrowty{k}{\tau}}% \tag{Borrows}\\
    \mid \svar{x}{\schm}^n % \tag{Suspended}
    % |&\ \tydecl{t}{\kschm}{K}{\tau}\tag{Type Declaration}\\
  \end{align*}
  \vspace{-10pt}
  \caption{Type environments}
  \label{grammar:env}
  \vspace{5pt}
  % \begin{minipage}{0.5\linewidth}
  \centering
    \input{sdtyping/split-binder}
    \vspace{-5pt}
    \caption{Splitting rules for bindings -- $\lsplit{C}{B}{B_l}{B_r}$}
    \label{sdtyp:split}
  \vspace{-10pt}
  % \end{minipage}
\end{figure}

@@

% Environments are also used to encode several linearity-related properties
% through \emph{environment-wide constraints}.
Constraints on an environment control substructural properties by
restricting the types of variables.  The constraint $\Cleq{\E}{k}$
abbreviates $\Cleq\schm k$, for all $\bvar{x}{\schm}$
% $\svar{x}\schm$, and $\bvar{\borrow{x}}{\borrowty{k'}{\schm}}$
in $\E$, which in turn means that $k' \lk k$ where $k'$ is the kind of
the variable's type scheme.
Borrow bindings follow the exact same rules, but suspended bindings
are forbidden in an environment $\E$ constrained like that.
This intuitive explanation is sufficient to understand
the {\sc Abs} and {\sc Var} rules shown in
\cref{selectrules:binders}.

@@

Elimination of pairs is done using a matching construct
$(\matchin{x,x'}{e_1}{e_2})$.
This construct is mostly standard, except it can operate both
on normal pairs and borrows of pairs.
The intuition is as follow:
A syntactic marker $\etransfm$ indicates if it applies to
a pair ($\etransfm = \operatorname{id}$) or a borrow ($\etransfm = \&^\BORROW$).
If $\etransfm = \operatorname{id}$, the typing simplifies to
a regular destruct on pairs.
Otherwise, $e_1$ is expected to be of type
$\borrowty{k}{\tyPair{\tau_1}{\tau'_1}}$
while $x$ and $x'$ are of type respectively
$\borrowty{k}{\tau_1}$ and $\borrowty{k}{\tau'_1}$.

@@

As a model of resource, we consider a simple abstract
type $\tapp{\tres}{\tau}$ whose content of type $\tau$ must be $\kun_0$ and
equipped with the four class of operations describe in \cref{motivation}:
\begin{itemize}[topsep=0pt]
\item
$\create$:
$\forall{\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\
\qual{\Cleq {\kvar_\tvar} {\kun_0}}
{\tvar \tarr{} \tapp\tres\tvar}$
% Creates a new resource.
\item
$\observe$:
$\forall{\kvar\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\
\qual{\Cleq {\kvar_\tvar} {\kun_0}}
{\borrowty[\IBORROW]{\kvar}{\tapp\tres\tvar} \tarr{} \tvar}$
% Reads from a shared borrow of a resource.
\item
$\update$:
$\forall{\kvar\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\
\qual{\Cleq {\kvar_\tvar} {\kun_0}}
\borrowty[\MBORROW]{\kvar}{\tapp\tres\tvar} \tarr{} \tvar \tarr{\kaff} \tunit$
% Writes to an exclusive borrow of a resource.
\item
$\destroy$:
$\forall{\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\
\qual{\Cleq {\kvar_\tvar} {\kun_0}}
\tapp\tres\tvar \tarr{} \tunit$
% Destroys a resource.
\end{itemize}

@@

\lstMakeShortInline[keepspaces,style=rule,basicstyle=\normalsize\normalfont]@

The dynamics of \lang is given in big-step
functional
style~\cite{siek13:_type_safet_three_easy_lemmas,DBLP:conf/esop/OwensMKT16,
  DBLP:conf/popl/AminR17}.  A function
@eval \Store \Perm \VEnv i e@
manipulates the semantic objects defined in
\cref{fig:syntax-internal-language}.
% To demonstrate resource handling, the semantics comes with an abstract
% notion of linear resources that can be created, destroyed, observed
% (through unrestricted borrows), and
% updated (through affine borrows).
%
The semantics is defined in terms of \emph{elaborated expressions} $e$
with kind, constraint, and splitting annotations inserted by the typechecker.
A splitting $\Sp$ is evidence of the splitting relation for type environments
used in the typing rules.

Let-polymorphism is demonstrated for functions giving rise to elaborated
$\eletfun$ expressions annotated with a type scheme $\schm$ and a kind $k$ indicating their
usage restriction (linear, affine, etc) relative to the variables and
constraints of $\schm$. Their use
gives rise to explicit instantiation of the kind and type variables.
Pairs come with a kind tag $k$ indicating the usage restriction.

@@

A store $\Store$ is a partial map from raw locations to
storables. The function
@salloc: store -> storable -> (loc * store)@ is such that
@salloc delta@ @w@ allocates an unused location in @delta@ and fills it with
@w@. It returns the location and the extended store.

The evaluation function is indexed by a step count @i@ so that each
invocation is guaranteed to terminate either with an error, a timeout,
or a result. Its return type is a monad
@'a sem@ which combines error reporting and timeout:
\begin{lstlisting}
type 'a sem = Error of string | TimeOut | Ok of 'a
val eval: store->perm->venv->int->exp->(store*perm*result) sem
\end{lstlisting}
It evaluates the given expression in the context of an initial store, a
permission to use addresses in the store, a value environment, and a
step count. If successful, it returns the final store, the remaining
permissions, and the actual result.

@@

We give some excerpts of the definition of @eval@ in
\cref{fig:big-step-interpretation} and leave the full
definition for \cref{sec:semant-defin}.
% The appendix also
% contains the semantics in mathematical notation.
The definition uses OCaml syntax with some pretty
printing. The pervasive @let*@ operator acts as monadic bind
for the @sem@ monad. The operator
@let*? : bool -> (unit -> 'a sem) -> 'a sem@
 converts a boolean
argument into success or failure in the monad.
 The function header of @eval@ checks
whether time is up and otherwise proceeds processing the expression.

>>>>

\lang is a minimal ML-like language with
let-polymorphism and abstract types. Its type system manages
linearity and borrowing using kinds and (kind-) constrained types.
For ease of presentation, we consider a simplified internal language.
\begin{itemize}[topsep=0pt]
\item Pattern matching is demonstrated on pairs rather than algebraic
  datatypes.
\item There are separate operators for borrowing and reborrowing (taking
  the borrow of a borrow); the surface language unifies these
  operators using ad-hoc polymorphism / typeclasses.
\item Regions are explicit in the code and must be annotated using the
  algorithms presented in \cref{regionannot}.
\item Regions are identified using nesting levels instead of region
  variables.
\end{itemize}

@@

A borrow $\borrow{x}$ is always taken from a variable $x$. The
\emph{borrow annotation} $\BORROW$ indicates whether the borrow is exclusive
($\MBORROW$) or shared ($\IBORROW$). If $x$ is already borrowed,
then we need to use the reborrow expression $\reborrow{x}$.
%
A region $\region{\Sone x\BORROW}{e}$ is annotated with its nesting $n$, the variable $x$ that may be borrowed in $e$, and the kind of borrow $\BORROW$.
%
A match is indexed with a \emph{match specification} $\etransfm$ that indicates
whether the match operates on borrows ($\etransfm=\&^b$) or not ($\etransfm=id$).
%
We consider four primitive operations to manipulate resources:
$\create$, $\observe$, $\update$ and $\destroy$.

@@

In the surface language (\cref{motivation}) region annotations are optional.
In the internal language, regions must be syntactically explicit and
annotated with a nesting index and a scoped variable.
This section defines a transformation $\RannotT{p}{p'}$ which
automatically inserts region annotations in  programs.
The input $p$ is a program with optional region annotations of the form $\region[]{}{e}$.
The output $p'$ is a program with explicit annotations
of the form $\region{\Sone x\BORROW}{e}$ such that no borrow occurs
outside a region.
We give an informal presentation of our code transformation and defer
the complete definition to \cref{appendix:regionannot}. This code transformation
aims to find, for each borrow $\borrow{x}$, the biggest region satisfying
the following rules:
\begin{enumerate}
\item The region should contain at least $\borrow{x}$.
\item The region must be contained in the scope of $x$.
\item An exclusive borrow $\borrow[\kaff]{x}$ should never share a region with any other borrow of $x$.
\item The variable $x$ cannot occur in the region of $\borrow{x}$.
\end{enumerate}
The transformation starts from each borrow $\borrow x$ and grows its associated
region until enlarging it would include the binding for $x$ or lead to a
conflicting use of $x$ or a conflicting borrow for $x$.
As an example, consider the following program:
%
\[
  \setlength{\jot}{-1pt}
  \RannotT{
\begin{aligned}
  \lam{a}{}\ &
  \letin{x}{\app{f}{\borrow[\kun]{a}}}{}\\
  &g\ (\borrow[\kaff]{x});\\
  &f\ (\borrow[\kun]{x})\ (\borrow[\kun]{x})\\
\end{aligned}
}{
\begin{aligned}
  \lam{a}{}\region[1]{\Sone a\MBORROW}{\ &%
  \letin{x}{\app{f}{\borrow[\kaff]{a}}}{}\\
  &\region[2]{\Sone x\MBORROW}{g\ (\borrow[\kaff]{x})};\\
  &\region[2]{\Sone x\IBORROW}{f\ (\borrow[\kun]{x})\ (\borrow[\kun]{x})}\\
  &\!\!\!}
\end{aligned}
}
\]
%
As variable $a$ only has one borrow, its region covers its whole lexical scope.
Variable $x$ has multiple conflicting borrows and requires more consideration.
We place a first
region around the exclusive borrow and its function application,
and a second region around both shared borrows. This placement of region
is optimal: making any region bigger would cause a borrowing
error.
Region indices are assigned after placement, so it is trivial to
ensure well-nested regions where the inner regions have
higher indices than the outer ones.

Programmers may also annotate regions explicitly.
The transformation considers an annotation as an
upper bound on the contained regions.
In the following program, a manual annotation has been inserted to ensure
no borrow enters the reference $r$:
%
\[
  \setlength{\jot}{-1pt}
  \RannotT{
\begin{aligned}
  \letin{&r}{\operatorname{ref}\ 0}{}\\
  \lam{a}{}\ &
  set\ r\ \region[]{}{g\ (\borrow[\kun]{a})};\\
  &f\ (\borrow[\kun]{a})
\end{aligned}
}{
\begin{aligned}
  \letin{&r}{\operatorname{ref}\ 0}{}\\
  \lam{a}{}\ &
  set\ r\ \region[1]{\Sone a\IBORROW}{g\ (\borrow[\kun]{a})};\\
  &\region[1]{\Sone a\IBORROW}{f\ (\borrow[\kun]{a})}
\end{aligned}
}
\]
%
The rules allow merging the two regions around the borrows
$\borrow[\kun]{a}$. However
the explicit annotation indicates that the region should stay around the closure
passed as argument of $set$. This feature is useful to control captures
by imperative APIs.

The code transformation is purely syntactic and must be used before
typing. It only produces well-nested annotations: if $\region[n]{b}{\dots}$
is nested inside $\region[n']{b'}{\dots}$, then $n > n'$. Furthermore, there
is at most one region per borrow, and exactly one region per exclusive borrow.
In the rest of this article, we assume
that all terms have been annotated by this code transformation and respect
these properties.

@@

To avoid distraction, this section focuses on the essential and novel
parts of the type system. A complete 
description is available in \cref{appendix:sdtyping}.
% In this section, we focus on chosen parts of our system
% through a set of selected rules.
%
Here we only discuss the following judgments:
\begin{flushleft}
  \begin{tabular}{l@{ --- }l}
    $\inferS{C}{\E}{e}{\tau}$&
                               Expression $e$ has type $\tau$ in
                               environment $\E$ under constraints
                        $C$. \\
    $\inferSK{C}{\E}{\tau}{k}$&
                                Type $\tau$ has kind $k$ in
                                environment $\E$ under constraints
                                $C$. \\
    $\entail{D}{C}$ & Constraint $D$ entails constraint $C$.
  \end{tabular}
\end{flushleft}

@@

\affe uses kinds and constrained types to indicate
linear and affine types.
% Before explaining the details of our type system, we
% give a quick presentation of the kinds and constraints.
A kind $k$ is either a kind variable, $\kvar$, or a constant $\mul_n$.
The quality $\mul$ describes the use pattern of the type:
unrestricted $\kun$, affine $\kaff$, or linear $\klin$. The level $n \in \Nat \cup \{\infty\}$
describes the nested regions in which the value can be used.
Level $0$ refers to the top-level scope outside any region; we often elide it
and write $\kaff$ for $\kaff_0$. Level
$\infty$ refers to an empty region that is infinitely nested.
For instance, the constraint $\Cleq{\kvar}{\kun_\infty}$ indicates that
$\kvar$ must be unrestricted, but can be local to a region.
%
Kinds form a lattice described in \cref{sdtyp:lattice}.
Unrestricted values may be used
where affine ones are expected and affine ones are less restricted
than linear ones as reflected in \TirName{Lat-UAL}.
Values usable at level $n$ may be used at any more deeply
nested level $n'$ as defined in the \TirName{Lat-Level} axioms.
%
Constraints are conjunctions of inequality constraints over this kind
lattice, i.e., they specify upper or lower bounds for
kind variables or relate two kind variables.

@@

\begin{figure}[tp]
  \begin{minipage}{0.36\linewidth}
  \begin{align*}
    \E ::=&\ \Eempty \mid \E;B\tag{Environments}\\
    B ::=&\ \bnone \tag{Empty} \\
    |&\ \bvar{\tvar}{\kschm} \tag{Types}\\
    |&\ \bvar{x}{\schm} \tag{Variables} \\
    |&\ \bvar{\borrow{x}}{\borrowty{k}{\tau}} \tag{Borrows}\\
    |&\ \svar{x}{\schm}^n \tag{Suspended}
    % |&\ \tydecl{t}{\kschm}{K}{\tau}\tag{Type Declaration}\\
  \end{align*}
  \vspace{-10pt}
  \caption{Type environments}
  \label{grammar:env}
  \end{minipage}\hfill
  \begin{minipage}{0.6\linewidth}
    \input{sdtyping/split-binder}
    \vspace{-5pt}
    \caption{Splitting rules for bindings -- $\lsplit{C}{B}{B_l}{B_r}$}
    \label{sdtyp:split}
  \end{minipage}
\end{figure}

@@

% Environments are also used to encode several linearity-related properties
% through \emph{environment-wide constraints}.
Constraints on an environment control substructural properties by
restricting the types of variables.  The constraint $\Cleq{\E}{k}$
stands for the conjunction of $\Cleq\schm k$, for all $\bvar{x}{\schm}$
% $\svar{x}\schm$, and $\bvar{\borrow{x}}{\borrowty{k'}{\schm}}$
in $\E$, which in turn means that $k' \lk k$ where $k'$ is the kind of
the variable's type scheme $\schm$.
Borrow bindings follow the same rules, but suspended bindings
are forbidden in an environment $\E$ constrained like that.
This intuitive explanation is sufficient to understand
the \TirName{Var} and \TirName{Abs} rules shown in
\cref{selectrules:binders}.

@@

Elimination of pairs is done using a matching construct
$(\matchin{x,x'}{e_1}{e_2})$.
This construct is mostly standard, except it can operate both
on a normal pair and a borrow of a pair.
The intuition is as follows:
A syntactic marker $\etransfm$ indicates if it applies to
a pair ($\etransfm = \operatorname{id}$) or a borrow ($\etransfm = \&^\BORROW$).
If $\etransfm = \operatorname{id}$, the typing simplifies to
the usual elimination of a pair.
Otherwise, $e_1$ is expected to be a borrow of type
$\borrowty{k}{\tyPair{\tau_1}{\tau'_1}}$
and the variables $x$ and $x'$ have type
$\borrowty{k}{\tau_1}$ and $\borrowty{k}{\tau'_1}$,
respectively. Thus, the borrow of a pair is considered as a pair of
borrows of its components.

@@

To demonstrate how \lang{} deals with resources, we introduce an abstract
type $\tapp{\tres}{\tau}$ whose content of type $\tau$ must be
unrestricted ($\kun_0$) and which is equipped with the four
operations described in \cref{motivation}: 
\begin{itemize}[topsep=0pt]
\item
$\create$:
$\forall{\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\
\qual{\Cleq {\kvar_\tvar} {\kun_0}}
{\tvar \tarr{} \tapp\tres\tvar}$
% Creates a new resource.
\item
$\observe$:
$\forall{\kvar\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\
\qual{\Cleq {\kvar_\tvar} {\kun_0}}
{\borrowty[\IBORROW]{\kvar}{\tapp\tres\tvar} \tarr{} \tvar}$
% Reads from a shared borrow of a resource.
\item
$\update$:
$\forall{\kvar\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\
\qual{\Cleq {\kvar_\tvar} {\kun_0}}
\borrowty[\MBORROW]{\kvar}{\tapp\tres\tvar} \tarr{} \tvar \tarr{\kaff} \tunit$
% Writes to an exclusive borrow of a resource.
\item
$\destroy$:
$\forall{\kvar_\tvar}\bvar{\tvar}{\kvar_\tvar}.\
\qual{\Cleq {\kvar_\tvar} {\kun_0}}
\tapp\tres\tvar \tarr{} \tunit$
% Destroys a resource.
\end{itemize}

@@

% PT: removed style=rule
\lstMakeShortInline[keepspaces,basicstyle=\normalsize\normalfont]@

The dynamics of \lang is given in big-step
functional
style~\cite{siek13:_type_safet_three_easy_lemmas,DBLP:conf/esop/OwensMKT16,
  DBLP:conf/popl/AminR17}.  A function
@eval@
manipulates the semantic objects defined in
\cref{fig:syntax-internal-language}.
% To demonstrate resource handling, the semantics comes with an abstract
% notion of linear resources that can be created, destroyed, observed
% (through unrestricted borrows), and
% updated (through affine borrows).
%
The semantics is defined in terms of \emph{elaborated expressions} $e$
with kind, constraint, and splitting annotations inserted by the typechecker.
A splitting $\Sp$ is evidence of the splitting relation for type environments
used in the typing rules.

Let-polymorphism in the surface language gives rise to elaborated
$\eletfun$ expressions annotated with a type scheme $\schm$ and a kind $k$ indicating their
usage restriction (linear, affine, etc) relative to the variables and
constraints of $\schm$. Their use
gives rise to explicit instantiation of the kind and type variables.
Pairs come with a kind tag $k$ indicating the usage restriction.

@@

A store $\Store$ is a partial map from raw locations to
storables. The function
@salloc: store -> storable -> (loc *store)@ is such that
@salloc delta@ @w@ allocates an unused location in @delta@ and fills it with
@w@. It returns the location and the extended store.

The evaluation function is indexed by a step count @i@ so that each
invocation is guaranteed to terminate either with an error, a timeout,
or a result. Its return type is a monad
@'a sem@ which combines error reporting and timeout:
\begin{lstlisting}
type 'a sem = Error of string | TimeOut | Ok of 'a
val eval: store->perm->venv->int->exp->(store *perm *result) sem
\end{lstlisting}
Function @eval@ evaluates the given expression in the context of an initial store, a
permission to use addresses in the store, a value environment, and a
step count. If successful, it returns the final store, the remaining
permissions, and the actual result.

@@

We give some excerpts of the definition of @eval@ in
\cref{fig:big-step-interpretation} and leave the full
definition for \cref{sec:semant-defin}.
% The appendix also
% contains the semantics in mathematical notation.
The definition uses OCaml syntax with extensive pretty
printing. The pervasive @let*@ operator acts as monadic bind
for the @sem@ monad. The operator
@let*? : bool -> (unit -> 'a sem) -> 'a sem@
 converts a boolean
argument into success or failure in the monad.
\begin{lstlisting}
let (let*?) : bool -> (unit -> 'b sem) -> 'b sem =
  fun b f -> if b then f () else Error ("test failed")
\end{lstlisting}
The function header of @eval@ checks
whether time is up and otherwise proceeds processing the expression.
